---
title: "3 - Build swarm space"
author: "Marina Papadopoulou"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3 - Build swarm space}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## 3.1 Load data

Load the metrics of collective motion calculated previously. Include the events you want to compare in a swarm space.

```{r message=FALSE, warning=FALSE}
library(swaRmverse)
data("multi_species_metrics")

all_data <- multi_species_metrics
## or bind with new data
#all_data <- rbind(multi_species_metrics,
#                  new_species_metrics)

```

## 3.2 Build swarm space

Create new swarm space with PCA analysis:

```{r}
new_pca <- create_swarm_space(metrics_data = all_data,
                              space_type = "pca"
                              )

ggplot2::ggplot(new_pca$swarm_space,
                 ggplot2::aes(x = PC1, y = PC2, color = species)
                ) +
                ggplot2::geom_point() +
                ggplot2::theme_bw()



```
Check what is principal component represents:

```{r}
pcs_info <- new_pca$pca$rotation[, new_pca$pca$sdev > 1]
print(pcs_info)

```

Or create a new swarm space with tSNE to better study the local structure of the data:

```{r}

new_tsne <- create_swarm_space(metrics_data = all_data,
                              space_type = "tsne",
                              tsne_rand_seed = 2023,
                              tsne_perplexity = 10
                              )

ggplot2::ggplot(new_tsne, ggplot2::aes(x = X, y = Y, color = species)) +
  ggplot2::geom_point() +
  ggplot2::theme_bw()

```

## 3.3 Expand existing swarm space

Starting from previously generated PCA swarm space, add new data:

```{r}
#load('../data/pca_res.RData')
# expanded_pca <- expand_pca_swarm_space(metrics_data = new_species_metrics,
#                                        pca_data = new_pca,
#                                        pca_space = pca_res)
#
# ggplot2::ggplot(expanded_pca,
#                 ggplot2::aes(x = PC1, y = PC2, color = species)) +
#    ggplot2::geom_point() +
#    ggplot2::theme_bw()
```
